Estimating duration in the presence of misclassification

### 2024-03-18 introduction.tex
first draft of conclusion
Updates from Overleaf

COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}
Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current mean impact of pandemics is over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

\section{Surveillance metrics}

@@

The most common reproduction number is the \emph{basic reproduction number}, $\R$.
$\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
A fully susceptible population means none of its members have any immunity to the disease.
Note that this definition implies $\R$ is constant for a given pathogen and population.
In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
$\R$ has a threshold value of 1: an outbreak has the potential to cause an epidemic if and only if $\R < 1$~\autocite[76]{diekmannMathematical}.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
Yet the size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
To translate incidence into the burden of the disease, a measure of severity is required.
For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.

\emph{Transmission} refers to the process by which an infectious disease generates new infections.
Estimating transmission refers to estimating the parameters of this process, the infection process, including its components.
The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
They are also important for forecasting future incidence.

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Whether a given individual is \emph{prevalent} (has the disease) is often measurable through a diagnostic test.
Therefore, the prevalence of the population can be estimated with a prevalence survey.
However, these are very expensive to power appropriately; SARS-CoV-2 is unusual in that its impact is so large that the UK government did this.
Prevalence and incidence are strongly related, with the relationship determined by by the duration of an infection (see \cref{E-inc-prev}).

@@

However, incidence and transmission and prevalence, are almost never directly observed.
Therefore, they need to be inferred from other data sources.

@@

\todo[inline]{Not sure this section (data sources) needs to exist, particularly in this location}
\todo[inline]{If this section kept, probably want to cite some things on generic data for pandemic surveillance}

The data sources used for outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify.
For example, community testing which relies on individuals presenting for a test, hence biased by the willingness of individuals to be tested.

Well-designed studies can provide direct estimates of incidence and/or prevalence in specific cohorts.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, for these studies to be appropriately powered, they often require a large cohort, and hence are expensive to run.
A unique feature of the COVID-19 pandemic is that its impact was so large that the UK Government were willing to fund such studies.
The availability of this data motivates the work in this thesis.

\section{Infection process model}

The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of how incidence arises.
A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process.
Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates, but can introduce bias and are often computationally challenging to fit.

The models with the weakest assumptions are statistical models.
These make almost no assumptions about the transmission process, viewing incidence as a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
The most common model is that incidence follows a Poisson process, and the intensity of the process is the target of incidence.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
The most common type of semi-mechanistic model are those based on \emph{renewal equation}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
\end{align}
where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
The effective reproduction number can be constructed as a function of $\beta(t, \tau)$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature futher in this thesis, for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex and aim to fully represent the transmission process.
\emph{Compartmental models} are the most common type of mechanistic model.
Compartmental models split the population into compartments, each representing a different state of the disease (described in full in \cref{E-SEIR:sec:mechanistic-models}).
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions, in particular the age structure of the population is important for respiratory diseases.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

@@

\todo[inline]{Needs some generic citations about observation models}

The \emph{observation model} links the transmission model to the data.
The choice of observation model can, in theory, be independent of the transmission model.
However, pragmatic considerations often mean that the observation model is chosen to be in a similar form to the transmission model.

Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
This equation links an observation at time $t$, $y_t$, to the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    y_t &= \zeta \sum_{i=0}^\infty I(t-i) f(i)
\end{align}
where $I(t)$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.

% The choice of an observation model must be heavily informed by the data that will be used.
% Two types of data commonly arise in the context of infectious disease: incidence measures and prevalence measures.

% An incidence measure is characterised by each infection only appearing in the dataset at most once.
% Common examples are hospital admissions and deaths.
% The basic form of an observation model for an incidence measure is a scaling and delay of the incidence.
% The scaling is the proportion of infections that are observed (\eg the proportion of infections that are fatal), and the delay is the time between an infection occurring and it being observed (\eg the time between infection and death).
% In discrete time this is:
% \begin{align}
%     M(t) = \zeta \sum_{\tau=0}^\infty I(t-tau) f(\tau)
% \end{align}
% where $M(t)$ is the observation of the incidence measure at time $t$; $I(t)$ is the incidence at time $t$; $\zeta$ is the scaling; and $f(\tau)$ is the delay, formalised as the probability that an infection is observed $\tau$ days after it occurs, given that it will be observed.

% A prevalence measure is characterised by each infection possibly appearing in the dataset multiple times.
% Common examples are the results of a prevalence survey or the number of individuals in hospital with a disease.
% \todo[inline]{Finish writing about prevalence measures}

\todo[inline]{Expand slightly on convolution equations and why solving them is challenging, including some citations}
Solving a convolution equation for $I$ is an ill-posed problem when $f(0)$ is small.
Either an approach which gives a large $f(0)$, for example by assuming that the delay between incidence and being prevelant is 0, or smoothing on $I(t)$ is required.

\section{Backcalculation}

\todo[inline]{
  Decide whether to keep this section.
  I think backcalculation probably does need to be called out explicitly, but it feels a little out-of-place.
  IF kept, needs expanding a bit, flow a bit better, and citations added.
}

\emph{Backcalculation} was a modelling framework developed in the context of HIV/AIDS but since applied more broadly.
The aim of backcalculation is to estimate the incidence of a disease from the observation of a prevalence measure.
It combines a statistical transmission model with an observation model for prevalence data.

Here, $\zeta = 1$ and $f(t)$ represents the probability of having the disease $t$ days after infection.
In this way, $f$ represents a duration.

\todo[inline]{Following paragraph seems a bit specific but is needed to motivate the following section}
For a prevalence survey, the delay distribution is the \emph{duration of positivity}, the distribution of the time between an individual becoming infected and the time they are no longer infected.

\section{Estimating the duration of positivity}

Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
Modelling a biomarker uses more information per observation, but requires stronger assumptions.

@@

Modelling the duration directly considers only the binary result at each test (positive or negative); for instance in a \emph{survival analysis} framework.
This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.

@@

\section{Application to SARS-CoV-2 in England}

\subsection{Data for estimating incidence and transmission} \label{intro:sec:data-for-estimating}

@@

For large parts of the pandemic, community testing was widely available in the UK.
However, it does not test a random sample of the population.
The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.

Seroprevalence surveys estimate the proportion of the population previously infected.
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representativeness of the wider population is unclear.

@@

Prevalence surveys in representative samples of the population are the most reliable measure of the burden of the disease.
They are not subject to the biases of community testing and observe the whole population, unlike severe events, although they can suffer from non-response bias.
However, they are expensive to power to inform detailed analyses.
Often, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).

The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
These surveys use RT-PCT (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).

@@

In this thesis, I make use of two cohort studies with differing designs to estimate duration.
These designs have complementary strengths, and I combine information from the studies.
Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.

@@

The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
In addition to estimating prevalence, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up mean that the tails of the distribution of the duration of positivity can be estimated.
However, the lack of frequent testing (at most weekly) means that the viral load dynamics cannot be estimated.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

\subsection{SPI-M-O}

In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.

@@

I will investigate whether estimates relying on only one data source, the CIS, could replace this process.
Such estimates should be more robust, due to the representative nature of CIS.

\subsection{Use of prevalence surveys}

@@

\subsection{Previous duration estimates}

@@

@@

A particular challenge with these analyses is that they must be performed within the SRS because they require individual-level data.
The SRS is a form of TRE (Trusted Research Environment), with limited computing power.
Analyses with long run times (\ie days) and/or requiring large amount of parallelisation (\eg a high-performance cluster) are not feasible.

>>>>

COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}
Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current mean impact of pandemics is over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

\section{Surveillance metrics} \label{intro:sec:metrics}

@@

The most common reproduction number is the \emph{basic reproduction number}, $\R$.
$\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
A fully susceptible population means none of its members have any immunity to the disease.
Note that this definition implies $\R$ is constant for a given pathogen and population.
In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
$\R$ has a threshold value of 1: an outbreak has the potential to cause an epidemic if and only if $\R > 1$~\autocite[76]{diekmannMathematical}.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
Yet the size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
To translate incidence into the burden of the disease, a measure of severity is required; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.

\emph{Transmission} refers to the process by which an infectious disease generates new infections.
Estimating transmission refers to inferring the parameters of this process, including its components.
The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
They are also important for forecasting future incidence.

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Whether a given individual is \emph{prevalent} (has the disease) is often measurable through a diagnostic test.
Therefore, the prevalence of the population can be estimated with a prevalence survey, although these are typically very expensive to power appropriately.
Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).

@@

Collecting data which directly informs inference of the quantities in this section is expensive or impossible.
Hence, they are typically inferred from other data sources through modelling.

@@

The data sources used for outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify.
Here I discuss three types of data which are useful for surveillance.
Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission, however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
These quantities can be estimated from other studies, \eg cohort studies.

\subsection{Incidence-type data}

Incidence-type data originate from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
This type of data is typically affected by \emph{ascertainment bias}: the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocite[chapter 9]{lashModern}.

An example of incidence-type data is community testing.
This relies on individuals presenting for a test, hence is biased by individuals' ability and willingness to be tested.

More severe symptoms are typically less affected by ascertainment bias.
An extreme example is the development of AIDS following infection by HIV.
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2).
This means that the observations do not provide much information on incidence in those subgroups.

\subsection{Prevalence-type data}

Prevalence-type data originate from observations of the number or proportion of a population in a specific disease state at a specific time.
For example, the number of hospital beds occupied with individuals with the disease or the proportion of the population with a marker of prior infection.
The issues associated with incidence-type data also apply to prevalence-type data.

Prevalence surveys are one form of prevalence-type data that can be unbiased.
Here, a representative proportion of the population (or a subpopulation of interest) are tested for a marker of current infection.
For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.

However, \emph{seroprevalence} surveys are more feasible.
Seroprevalence is the proportion of individuals in the population which have antibodies to the disease.
The presence of antibodies indicates that the individual was previously infected.
Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply.
However, seroprevalence is a delayed measure because there is a lag between an individual being infected and them developing antibodies.

Other epidemiological and/or biological quantities are normally required to estimate transmssion.
These depend on the model type, but could include the \emph{generation interval} (the time between an individual being infected and going on to infect further individuals) or duration of different disease stages.
Cohort studies, where a group of individuals are followed over time, are a study design which can estimate many of these quantities.

Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, for these studies to be appropriately powered, they often require a large cohort, and hence are expensive to run.

\section{Infection models}

The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which incidence arises.
A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process.
Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates, but can introduce bias and are often computationally challenging to fit.

The models with the weakest assumptions are statistical models.
These make almost no assumptions about the transmission process, viewing incidence as a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
The most common model is that incidence follows a Poisson process, and the intensity of the process is the target of inference.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
The most common type of semi-mechanistic model are those based on a \emph{renewal equation}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
\end{align}
where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
The effective reproduction number can be constructed as a function of $\beta(t, \tau)$ and the generation interval.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions, in particular incorporating heterogeneity; the age structure of the population is important for respiratory diseases.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

\section{Disease models}

The \emph{disease model} links the transmission model to a population-level quantity which we will make observations of.
Examples are the incidence or prevalence of symptomatic individuals and admissions to hospital.

The choice of observation model can, in theory, be independent of the transmission model.
However, pragmatic considerations often mean that the observation model is chosen to be in a similar form to the transmission model.

Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
This equation describes the expectation of the quantity at time $t$, $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.
$f$ is known as a \emph{delay distribution}.

Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem when $f(0)$ is small.
The forwards error of the solution is large when the ratio of the off-diagonal to diagonal elements is large~\autocite[chapter 8.2]{highamAccuracy}; for this situation, that is when the value of $\max_{i > 0} f(i) / f(0)$ is large.
A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}.
Either an approach which gives a large $f(0)$ or smoothing on $Z_t$ is required.

@@

An observation model links $W_t$ to a possibly biased and/or noisy observation of it at time $t$.
Biases can arise in a variety of ways, \eg healthcare-seeking behaviour or tests which are less than 100\% accurate.
Bias can be modelled through additional parameters or data quantifying these issues.
Noise could come from a variety of sources, \eg sampling error, clustering, or unexplained overdispersion.
Noise is typically modelled by a parameteric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption.

\section{Delay distributions}

Knowledge of a delay distribution, $f$ in \cref{intro:eq:disease-model}, is required for many surveillance procedures.
Typically, this is either the time until an event or the duration of a disease stage.

Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
Modelling a biomarker uses more information per observation, but requires stronger assumptions.

@@

Modelling the duration directly considers only the binary result at each test (positive or negative); for instance in a \emph{survival analysis} framework.
This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
It is also easier to specify non-parametric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.

@@

\section{SARS-CoV-2 in England}

The surveillance systems used in response to SARS-CoV-2 are historically unique.
Modern technology allowed large amounts of data to be collected, distributed, and analysed.
While the HIV/AIDS pandemic had inspired a large amount of statistical methodology to be developed, the pandemic was far slower moving.
Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.

Deep understanding of each data source's potential was not possible.
In this thesis, I will investigate the potential of the prevalence surveys in greater detail.

\subsection{Data sources} \label{intro:sec:data-for-estimating}

@@

\subsubsection{Incidence-type data}

For large parts of the pandemic, community testing was widely available in the UK.
However, it did not test a random or even representative sample of the population.
The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.

@@

\subsubsection{Prevalence-type data}

Seroprevalence surveys were conducted throughout the epidemic, primarly in blood donors.
In addition to the normal problem of the delay between infection and antibody response, blood donors are a  convenience samples whose representativeness of the wider population is unclear.

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were not subject to the biases of community testing and observed the whole population, unlike severe events, although suffered from non-response bias.
Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
Despite this issue, they were considered the gold-standard metric.

The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal study, testing a large number of individuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).

@@

\subsubsection{Cohort studies}

Many cohort studies to estiamte a wide variety of quantities ran throughout the pandemic.
In this thesis, I make use of two cohort studies with differing designs to estimate the duration or PCR positivity.
These designs have complementary strengths, and I combine information from the studies.
Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.

@@

The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up mean that the tails of the distribution of the duration of positivity can be estimated.
However, the lack of frequent testing (at most weekly) means that the viral load dynamics cannot be estimated.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

Early on in the epidemic, incidence could be estimated directly from the longitudinal survey data by considering the proportion of the cohort that tested negative then positive at consecutive tests~\autocite{onsIncidenceOld}.
However, the method was abandoned in autumn 2020 when the testing was too infrequent to ensure that all infections were captured (\ie tested positive at least once).

\subsection{Models}

In the UK, the most modelling was produced by SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group).
SPI-M-O was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
This used a wide range of  modelling approaches depending on the data available, which fluctuated throughout the epidemic, and the question being addressed.
Some were fit to data while others were purely used for simulation based on plausible parameter values.

@@

Some of these models were used to explore policy scenarios.
For example, each of the stages of lifting the lockdown in 2021 was supported by modelling studies informing the government's decision~\autocite{sageEvidence}.
These aided decision-makers in understanding plausible impacts of their actions.

\section{Thesis context}

In this thesis, I consider how prevalence surveys could be used for epidemic surveillance.
These surveys have well-understood designs which means that estimates derived from them are more robust.
A crucial quantity in their use will be the distribution of the duration of RT-PCR positivity.

There is a large amount of related work in this area, from both prior to the pandemic and during its course.

\subsection{Backcalculation}

\emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod}, but applied very broadly.
It specifies a particular combination of infection, disease, and observation model to relate times of HIV infection with times of AIDS diagnosis via the incubation period.
Further details are in the recent review \textcite{sunModeling}.
The theory behind the framework underpins much of the work in this thesis.

The original formulation used a Poisson process as the infection model; a disease model where all infections are eventually observed ($\zeta = 1$) and the delay distribution, $f$, is the incubation period; and an observation model where all diagnoses are observed, and hence has no influence on the data.

To apply backcalculation to SARS-CoV-2 prevalence surveys, adaptations to the disease and observation models are needed (elaborated on in \cref{E-inc-prev}).
The disease model must consider that this is now prevalence-type, rather than incidence-type data.
This difference induces a dependency structure between the observations on each day.
The observation model must consider that this is now a survey, with sampling error, rather than assuming all infections are observed.

\subsection{Existing duration estimates} \label{intro:sec:previous-duration-estimates}

\subsection{Existing estimation from prevalence surveys}

@@

\subsection{Secure research service}

The data from the CIS is stored in the SRS (Secure Research Service)~\autocite{onsSRS}.
The SRS gives accredited researchers working on projects in the public interest access to granular data which cannot be legally shared publicly.
Only non-disclosive data can be extracted from the SRS.
In particular, counts normally need to be greater than 5.

The SRS contains limited computational power.
It has\todo{insert SRS computational power}.
However, these are shared computational resources.
Due consideration must be given to other users.
Processes which run for multiple days are liable to be cancelled.

@@

A particular challenge with these analyses is that they must be performed within the SRS because they require individual-level data.

@@
